---
title: "PumpItUp EDA"
output: html_document
---

```{r}
library(tidyverse)
library(gridExtra)
library(rpart)
library(rpart.plot)
library(caret)
library(magrittr)
library(purrr)

options(scipen = 9999)

train <- read.csv("/Users/jasonzivkovic/Documents/PumpItUp_DrivenDataComp/DataFiles/training_set_values.csv", stringsAsFactors = FALSE, na.strings = c("NA", ""))
train_labels <- read.csv("/Users/jasonzivkovic/Documents/PumpItUp_DrivenDataComp/DataFiles/training_set_labels.csv", na.strings = c("NA", ""))

test <- read.csv("/Users/jasonzivkovic/Documents/PumpItUp_DrivenDataComp/DataFiles/test_set_values.csv", na.strings = c("NA", ""))
```



```{r}
train_labels_y <- train_labels$status_group

train <- train %>%
  left_join(train_labels, by = "id")

```

# Inspect the data

```{r}
print(colSums(is.na(train)))

str(train)

summary(train)

varlist <- function (df=NULL,type=c("numeric","factor","character"), exclude=NULL) {
   vars <- character(0)
   if (any(type %in% "numeric")) {
     vars <- c(vars,names(df)[sapply(df,is.numeric)])
   }
   if (any(type %in% "factor")) {
     vars <- c(vars,names(df)[sapply(df,is.factor)])
   }  
   if (any(type %in% "character")) {
     vars <- c(vars,names(df)[sapply(df,is.character)])
   }  
}

chars <- varlist(df = train, type = "character")

train %<>% mutate_at(chars, funs(factor(.)))

# Howe many levels are there in the character/factor variables:
train %>% 
  map(levels) %>% map(length)

# the 'recorded_by' variable has only one level, can remove it from both training and test as it won't be used in the final model
train$recorded_by <- NULL
test$recorded_by <- NULL


```


## Training Data Outcomes

```{r}
train %>% group_by(quantity, quantity_group) %>% tally()

ggplot(data = train, aes(x=status_group, fill = status_group))+
  geom_bar(stat = "count") +
  labs(x= "Status Group", y="") +
  theme_bw()
```

# EDA of categorical variables

```{r}
# function to plot categorical variables

# cat_plots <- function(cols, df){
#   for (col in cols) {
#     order.cols <- names(sort(table(train[,col]), decreasing = TRUE))
#   
#     num.plot <- qplot(df[,col]) +
#       geom_bar(fill = 'cornflowerblue') +
#       geom_text(aes(label = ..count..), stat='count', vjust=-0.5) +
#       theme_minimal() +
#       scale_y_continuous(limits = c(0,max(table(df[,col]))*1.1)) +
#       scale_x_discrete(limits = order.cols) +
#       xlab(col) +
#       theme(axis.text.x = element_text(angle = 30, size=12))
#   
#     print(num.plot)
#   }
# }
# 
# cat_plots('quantity', train)


```


```{r}
cp1 <- train %>%
  group_by(funder, status_group) %>%
  summarise(n=n()) %>%
  arrange(desc(n)) %>%
  head(n=15) %>%
  ggplot(aes(x=funder, y=n, fill = status_group)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~ status_group)
```

## Working out which features can be dropped for model building
Some of the features appear to be duplicates, or extremely similar. To simplify my model, I will determine which of these can be dropped.

### Waterpoint Variables

```{r, fig.width=10}

wp1 <- ggplot(data = train, aes(x=waterpoint_type, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

wp2 <- ggplot(data = train, aes(x=waterpoint_type_group, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

grid.arrange(wp1, wp2, ncol = 2)

# Can remove waterpoint_type_group variable - waterpoint_type shows there is a difference in the association between "communal standpipe" and "communal standpipe multiple".

train$waterpoint_type_group <- NULL
test$waterpoint_type_group <- NULL
```

### Source Variables

```{r, fig.width=10}
source1 <- ggplot(data = train, aes(x=source, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

source2 <- ggplot(data = train, aes(x=source_type, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

grid.arrange(source1, source2, ncol = 2)

# source will be used, can drop source_type; over 75% of source = lake are non-functional, whereas source = river is only ~30%
train$source_type <- NULL
test$source_type <- NULL

```

### Quantity Variables

```{r, fig.width=10}
qty1 <- ggplot(data = train, aes(x=quantity, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

qty2 <- ggplot(data = train, aes(x=quantity_group, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

grid.arrange(qty1, qty2, ncol = 2)

# Identical, will drop quantity_group
train$quantity_group <- NULL
test$quantity_group <- NULL


```

### Water Quality Variables

```{r, fig.width=10}
water1 <- ggplot(data = train, aes(x=water_quality, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

water2 <- ggplot(data = train, aes(x=quality_group, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

grid.arrange(water1, water2, ncol = 2)

# Water quality has additional levels that appear to differ in the status of the well so I will drop quality group
train$quality_group <- NULL
test$quality_group <- NULL

```

### Payment Variables

```{r, fig.width=10}
pay1 <- ggplot(data = train, aes(x=payment, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

pay2 <- ggplot(data = train, aes(x=payment_type, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

grid.arrange(pay1, pay2, ncol = 2)

# payment and payment_type the same - can drop one.
train$payment_type <- NULL
test$payment_type <- NULL

```

### Management Variables

```{r, fig.width=10}
man1 <- ggplot(data = train, aes(x=management, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

man2 <- ggplot(data = train, aes(x=management_group, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

grid.arrange(man1, man2, ncol = 2)

# I will leave both in for now, but may change this up.


```

### Extraction Variables

```{r, fig.width=10}
ext1 <- ggplot(data = train, aes(x=extraction_type, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

ext2 <- ggplot(data = train, aes(x=extraction_type_group, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

ext3 <- ggplot(data = train, aes(x=extraction_type_class, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

grid.arrange(ext1, ext2, ext3, ncol = 3)

# Again, I will go granular and only keep 'extraction_type'
train$extraction_type_group <- NULL
train$extraction_type_class <- NULL

test$extraction_type_group <- NULL
test$extraction_type_class <- NULL


```

### Construction Year

```{r}
train %>%
  #filter(construction_year != 0) %>%
  ggplot(aes(x=construction_year, fill = status_group)) +
  geom_density(alpha = 0.3)

train %>%
  filter(construction_year != 0) %>%
  ggplot(aes(x=construction_year, fill = status_group)) +
  geom_density(alpha = 0.3)

train$construction_year_bin <- cut(train$construction_year, seq(1960, 2020, 10)) %>% as.character()
train$construction_year_bin[is.na(train$construction_year_bin)] <- "Unknown"
train$construction_year_bin <- as.factor(train$construction_year_bin)

table(train$construction_year_bin)

test$construction_year_bin <- cut(test$construction_year, seq(1960, 2020, 10)) %>% as.character()
test$construction_year_bin[is.na(test$construction_year_bin)] <- "Unknown"
test$construction_year_bin <- as.factor(test$construction_year_bin)

table(test$construction_year_bin)

```


### Permit

```{r}
ggplot(data = train, aes(x=permit, fill = status_group)) +
  geom_bar(position = "fill") +
  coord_flip()

# will drop permit out 
train$permit <- NULL
test$permit <- NULL

```

```{r}
train %>%
  group_by(scheme_management) %>%
  summarise(n = n()) %>%
  arrange(desc(n))

# I will include scheme_management in the model, but exclude scheme_name. Will also replace NAs with "unknown"
train$scheme_management <- as.character(train$scheme_management)
train$scheme_management[is.na(train$scheme_management)] <- "Unknown"
train$scheme_management <- as.factor(train$scheme_management)

test$scheme_management <- as.character(test$scheme_management)
test$scheme_management[is.na(test$scheme_management)] <- "Unknown"
test$scheme_management <- as.factor(test$scheme_management)

train$scheme_name <- NULL
test$scheme_name <- NULL


```

```{r}
train %>% group_by(public_meeting) %>% tally()


train$public_meeting <- as.character(train$public_meeting)
train$public_meeting[is.na(train$public_meeting)] <- "Unknown"
train$public_meeting <- as.factor(train$public_meeting)

test$public_meeting <- as.character(test$public_meeting)
test$public_meeting[is.na(test$public_meeting)] <- "Unknown"
test$public_meeting <- as.factor(test$public_meeting)

```


```{r}
train %>%
  filter(population != 0) %>%
  ggplot(aes(x=population, fill = status_group)) +
  geom_density(alpha = 0.3)

train$population_bin <- cut(train$population, c(0,10,20,50,100,100000), include.lowest = TRUE)
table(train$population_bin)

train$population <- NULL

test$population_bin <- cut(test$population, c(0,10,20,50,100,100000), include.lowest = TRUE)

test$population <- NULL
```


```{r}
train %>% group_by(basin) %>% tally()

# will us basin for my geographical variable, and remove all other goegraphy based variables

train$longitude <- NULL
train$latitude <- NULL
train$subvillage <- NULL
train$region <- NULL
train$region_code <-NULL
train$district_code <- NULL
train$lga <- NULL
train$ward <- NULL

test$longitude <- NULL
test$latitude <- NULL
test$subvillage <- NULL
test$region <- NULL
test$region_code <-NULL
test$district_code <- NULL
test$lga <- NULL
test$ward <- NULL
  
train$num_private <- NULL
test$num_private <- NULL
```


```{r}
train$wpt_name <- NULL
train$installer <- NULL
train$date_recorded <- NULL

test$wpt_name <- NULL
test$installer <- NULL
test$date_recorded <- NULL

```


```{r}
train %>% group_by(funder) %>% tally() %>% arrange(desc(n))

train$funder <- NULL
test$funder <- NULL

```






